Brief Announcement: Tight Bounds for Online Balanced Repartitioning

### 2020-03-03 paper/paper.tex
Fork of the PODC submission

\title{Brief Announcement: An Improved Lower Bound for Dynamic Balanced Graph Partitioning}

\begin{comment}
\author{Maciej Pacut}
\email{maciej.pacut@univie.ac.at}
\orcid{0000-0002-6379-1490}
\affiliation{%
  \institution{Faculty of Computer Science, University of Vienna}
  \country{Austria}
}

@@

\keywords{online algorithms, competitive analysis, graph partitioning, clustering}

@@

\begin{abstract}
  Distributed   applications,  including  batch  processing, streaming, and scale-out databases generate a significant amount of network traffic and a~considerable fraction of their runtime is due to network activity. In this  paper, we study  algorithms  for collocating frequently communicating nodes in a distributed networked systems in an online fashion.

  In the online graph repartitioning problem, we are given a sequence of pairwise communication requests between n nodes, revealed online.
  The objective is to service these requests by partitioning the nodes into $\ell$ clusters, each of size $k$, such that frequently communicating nodes are located in the same cluster.
  The partitioning can be updated by migrating nodes between clusters for a~fixed cost.
  The goal is to jointly minimize the amount of inter-cluster communication and migration cost.

  So far, the best known lower bound for the competitive ratio of
	deterministic algorithms was $\Omega(k)$,
 by reduction from online paging.
  We improve this result significantly by providing the lower bound of $\Omega(k\cdot \ell)$.
  We follow up with an optimal algorithm for $k=3$.
  Finally,
  we provide a $O(k\cdot \ell)$-competitive algorithm for the variant
  where the optimal strategy is to move to a~fixed configuration
  that incurs zero communication cost through the whole request sequence.
\end{abstract}

@@

Main technical contribution of this paper is the lower bound $\Omega(k\cdot\ell)$ for the competitive ratio of any deterministic algorithm for online balanced repartition problem (see Section~\ref{sec:lowerbound}).
The best known lower bound so far was $\Omega(k)$ \cite{repartition-disc}
which involves only two clusters.

So far, the competitive ratio of the best known algorithm was the component-based algorithm with the ratio $O(k^2\cdot \ell^2)$~\cite{repartition-disc}.
By applying classic rent-or-buy technique, we obtained an optimal algorithm for $k=3$, that we present in Section~\ref{sec:k3}.
In Section~\ref{sec:ppl}, we study a restricted model of balanced repartition, called perfect partition.
For that, we present an~$O(k\cdot \ell)$-competitive algorithm, which is optimal.
We believe that this result is an important building block on the way to an optimal algorithm for the general model.
\mahmoud{The mention of distributed alg, very much out of context!} 
A distributed adaptation of a similar algorithm has been analyzed in \cite{sigmetrics19_partitioning}.

\section{Related Work}

The problem was studied under resource augmentation \cite{repartition-disc}, and under stochastic assumptions for the input\cite{stochastic-ring}.
For clusters of size $2$ (i.e., $k=2$), a constant competitive algorithm exists~\cite{repartition-disc}.

The model is related to online
caching~\cite{SleTar85,FKLMSY91,McGSle91,AcChNo00}. In our
setting, requests can be served remotely, which is
reminiscent of caching models \emph{with
bypassing}~\cite{EpImLN11,EpImLN15,Irani02}.

The static offline version of~the~problem, called the
\emph{$\ell$-balanced graph partitioning problem} is 
NP-complete, and cannot even be approximated within any finite factor unless P
= NP~\cite{AndRae06}. The static
variant where $\ell = 2$ corresponds to the minimum bisection problem, which
is already NP-hard~\cite{GaJoSt76}, and 
the currently best approximation ratio is $O(\log n)$~\cite{SarVaz95,ArKaKa99,FeKrNi00,FeiKra02,KraFei06,Raec08}.

\section{Preliminaries}
\label{sec:problem-definition}

The online \emph{balanced repartitioning} problem (\OBRP{}) is defined as
follows. We are given a set of $n$ nodes,
initially arbitrarily partitioned into $\ell$~clusters,
each of size~$k$.
The input consists of a sequence of pairwise communication requests
$\sigma = (u_1,v_1),$ $(u_2,v_2),$ $(u_3,v_3), \ldots$,
where a pair $(u_t,v_t)$ indicates that nodes $u_t$ and $v_t$ exchange a~fixed amount of data.
Nodes in $C \subset V$ are \emph{collocated}
if they reside in the same cluster.
An algorithm serves a communication request between two nodes
either \emph{locally} at cost~0
if they are collocated,
or \emph{remotely} at cost~1
if they are located in different clusters.
We refer to these two types of requests as \emph{internal}
and \emph{external} requests, respectively.
Before serving a request,
an online algorithm may \emph{repartition},
%(i.e., \emph{reconfigure}).
i.e.,
it may (optionally) move some nodes into clusters different than their current clusters while respecting the capacity of every cluster. 
Only after a repartitioning is completed,
the algorithm serves the  request.
The cost of migrating a node from one cluster to another
is~$\alpha \in \mathbb{Z}^+$.
For any algorithm $\ALG$,
its cost,
denoted by $\ALG(\sigma)$,
is the total cost of communications and
the cost of migrations performed by $\ALG$ while serving the sequence $\sigma$.
The objective is to minimize the total incurred cost.

In this paper, we use the technique of maintaining \emph{components} of nodes, similar to previous approaches~\cite{repartition-disc}. A~\emph{component} is a subset of frequently communicating nodes.
Initially,
every node is trivially a \emph{singleton} component.
An~algorithm is \emph{component respecting}
if it always keeps nodes that  belong to the same component collocated.
That is,
if the algorithm needs to move a node,
it moves the whole component containing the node.
%We say that the request $(u,v)$ is \emph{external}, if the algorithm is in the configuration, where $u$ and $v$ are located at different clusters.
%This is achieved by opting a sequence of  partitions (a.k.a.~configurations)
%such that frequently communicating nodes (from different clusters)
%are relocated to the same cluster before they inflict too much communication cost.
%A \emph{reconfiguration} (i.e.,\emph{ re-partitioning})
%is performed by migrating nodes between clusters,
%at the total cost of individual node migrations.
%The objective is to jointly minimize communication and migration costs while serving the sequence.

\section{Lower bound} %$\Omega(k\cdot \ell)$ for Competitive Ratio of Any Deterministic Algorithm}
\label{sec:lowerbound}

In this section, we provide a lower bound $\Omega(k\cdot \ell)$ for the competitive ratio of any deterministic algorithm.
It is sufficient to have clusters of capacity $k\geq 3$ for this lower bound.
For $k=2$ a constant competitive algorithm exists.

The adversary constructs components of nodes, and issues only requests among nodes in the same component.
If an online algorithm splits the components at any moment, the adversary issues requests to split pairs of nodes.
We utilize the fact that the online algorithm does not know the components of the adversary, and it may split some components while evicting nodes.
On the other hand, $\OPT$ never splits the components.

@@

\begin{theorem}
	The competitive ratio of any deterministic algorithm for \OBRP{} is in $\Omega(k\cdot \ell)$ for any $k\geq 3$.
\end{theorem}

\begin{proof}
	We construct an instance of the problem with $\ell$ clusters 
	$\set{ S_1, S_2,\dots , S_{\ell}}, |S_i|  = k$.
	Let $I(C)$ denote the cluster where nodes of a component $C$ are located initially.
	%	\maciek{So far components were not introduced; the best place to introduce these would be around the definition of \PPOBRP}.
	Fix any online algorithm \ALG{}.
	We construct the input sequence for \ALG{} as follows.
	First,
	we issue $k-2$ (internal) requests so that \ALG{} can form a component of $k-1$
	nodes on the cluster $S_1$.
	Recall that internal requests does not incur any cost for \ALG{}.
	If \ALG{} at any point splits a component
	(i.e., spreads its nodes over two or more clusters),
	then we continue to issue requests between non-collocated nodes of the component until \ALG{} collocates the nodes of the component.
	Hence, if \ALG{} never collocates them then it is not competitive.

	Let $x_0$  be the only single node left on $S_1$ and  $y_0 \in S_2$ be any single node on $S_2$.
	Next,
	we issue a request between $x_0$ and $y_0$.
	Since this is an external request,
	\ALG{} joins them into one component and collocates them in some cluster other than $S_1$.
	For this,
	\ALG{} moves to a new configuration
	that replaces $x_0$ and $y_0$ with two other single nodes $x_1$ and $y_1$ respectively.

	Next,
	we issue a request between $x_1$ and the largest component $C$ s.t.~$I(C) = I(x_1)$.
	\ALG{} must collocate $x_1$ and $C$ on some cluster other than $S_1$ and
	consequently replaces $x_1$ with some other (single) node $x_2$.
	We repeat issuing requests between the single node $x_i$ on $S_1$ and the largest component $C'$ s.t.~$I(C')=I(x_i)$,
  until there are only two single nodes left that  originate from the same cluster.
	Formally, we denote these two single nodes by $x^*, y^*$, and we have $I(x^*) = I(y^*)$, and~for any other pair of single nodes
	$x'$ and $y'$,
	we have $I(x') \neq I(y')$.
	At this point there are at most $\ell+1$ single nodes left,
	otherwise there would be more pairs of single nodes that were initially in the same cluster.

	Given this sequence of requests,
	the optimal strategy is to migrate $\set{x_0,y_0}$ to the cluster $I(x^*)$ by
	swapping $\set{x_0,y_0}$ with $\set{x^*,y^*}$.
	Hence,
	OPT pays for $4$ node migrations and
	\ALG{} incurs at least one migration for each node in the sequence $X := x_0, x_1,\dots$.
	We exclude at most $(k-1) + ( \ell+1)$ nodes out of $k \cdot \ell$ nodes,
	therefore $|X| \geq k \cdot \ell - k - \ell \in \Omega(k\cdot\ell)$.
\end{proof}

\section{An Optimal Algorithm for Clusters of Size 3}
\label{sec:k3}

In this section we show that the classic \emph{rent-or-buy} approach\cite{borodin-book, karlin-ski-rental} allows to obtain an optimal algorithm for $k=3$.
Upon receiving $\alpha$ requests between a pair of nodes, we collocate them in one cluster until the end of a phase (that we define precisely later).

%We define the algorithm \TAlg in the following way.
Now we describe the algorithm \TAlg.
It partitions nodes into components, and
initially each node belongs to its own singleton component.
For each pair of nodes, \TAlg maintains a counter, initiated to $0$. 
Upon receiving a request to a pair that is not collocated in one cluster, it increases their counter by $1$.
If the counter for a pair $(u,v)$ reaches $\alpha$, \TAlg merges the components of $u$ and $v$, and moves to the closest component respecting partitioning.
If no such partitioning exists, \TAlg resets all components to singletons, resets all counters to $0$, and ends the phase.

@@

In our analysis, we distinguish among three types of clusters: $C_1, C_2, C_3$. In a cluster of type $C_i$, the size of the largest component contained in this cluster is $i$.

\begin{lemma}
  \label{lem:1req}
  A single repartition of \TAlg consists of at most $2$ migrations.
\end{lemma}

\begin{proof}
  Observe that when the repartition is triggered by \TAlg, the resulting partitioning is component respecting.
  Otherwise, if it does not exist, \TAlg simply ends the phase and performs no repartition.

    Consider a request between $u$ and $v$ that triggered the repartition and let $U$ and $V$ be their respective clusters.
    Note that $U\neq V$,
	 as otherwise the request would not trigger a~repartitioning.
	 We consider cases upon types of clusters $U$ and $V$.
    Note that this is impossible that either $U$ or $V$ is of type $C_3$, as otherwise we merge components of size $3$, and no component respecting partitioning exists, a contradiction.
    If either $U$ or $V$ is of type $C_1$, then this cluster can fit the merged component, and the reconfiguration is local within $U$ and $V$.
    For two clusters, any configuration can be reached within two migrations, due to the fact that clusters are indistiguishable.

    Finally, we consider the case where both $U$ and $V$ are of type $C_2$. Note that $(u,v)$ cannot both belong to a component of size $2$, as this would mean that \TAlg has the component of size $4$, a contradiction with the case assumption that the component respecting partitioning exists. 
    Otherwise, if either $u$ or $v$ belongs to a component of size $2$, then it suffices to exchange components of size $1$ between $U$ and $V$.
    Finally, if $u$ and $v$ belong to components of size $1$, then we must place them in a cluster different from $U$ and $V$.
    Note that in such case, a $C_1$-type cluster $W$ exists, as otherwise no component respecting partitioning exists. In this case \TAlg performs one migration, i.e., it exchanges the nodes $u$ and $v$ with any two nodes of $W$.
\end{proof}

\pagebreak

@@

  As \TAlg is component respecting, it never increases any counter above $\alpha$.
  We say that the pair $(u, v)$ is \emph{saturated}, if the counter has value $\alpha$, and \emph{unsaturated} otherwise.

  Let $\sigma$ be the input sequence that arrived during the phase.
  Let $\sigma_{cost}$ be the requests that at the moment of arrival were external requests for \TAlg (these are the only requests that incurred a cost for \TAlg).
  In our analysis, we partition $\sigma_{cost}$ into subsequences $\sigma_I$ and $\sigma_E$.
  The sequence $\sigma_I$ (inter-component requests) are the requests from $\sigma_{cost}$ issued to pairs that belong to the same component of \TAlg at the end of the phase.
  The sequence $\sigma_E$ (extra-component requests) denotes the requests from $\sigma_{cost}$ that do not appear in $\sigma_I$.

  %Let $A^I$ be the cost of (extra-cluster) communication incurred in this phase by \TAlg between pairs that belonged to the single component at the end of the phase.
  %Let $A^E$ be the cost of (extra-cluster) communication incurred in this phase between the nodes that belong to different components at the end of this phase.
  Let $\TAlg(M)$ be the cost of migrations performed by \TAlg in this phase.
  \TAlg performs at most $2 \ell$ component merge operations, as
  exceeding this number means that a component of size $4$ exists, and the phase should have ended already.
  Combining this with Lemma~\ref{lem:1req} gives us $\TAlg(M) \leq 4\alpha\cdot\ell$.
  %Together with Lemma~\ref{lem:1req}, this allows us to bound the cost of migrations, $\TAlg(M) \leq 6\cdot\alpha\cdot\ell$.

  Now we bound $\TAlg(\sigma_I)$.
  Cluster of type $C_3$ contributes at most $3 \alpha - 1$ to $\TAlg(\sigma_I)$, as $2$ of pairs of nodes from the component are saturated and contribute $\alpha$ each, and the third, unsaturated pair contributes at most $\alpha-1$.
  Other cluster types contribute less: $C_1$ contributes $0$ and $C_2$ contributes $\alpha$.
  Summing this over all $\ell$ clusters gives us $\TAlg(\sigma_I) \leq (3 \alpha-1)\cdot \ell \leq 3\alpha\cdot\ell$.

@@

  Now we estimate the cost of $\OPT(\sigma_E)$ while accounting savings from migrations.
  By performing a swap of nodes $(u,v)$, $\OPT$ collocates $u$ with two nodes $u', u''$, and $v$ with two nodes $v'$, $v''$.
  This may allow to serve requests between $(u,u')$, $(u,u'')$, $(v,v')$ and $(v,v'')$ for free afterwards.
  As $\sigma_E$ consists of requests to unsaturated pairs, and it accounts only for external requests, there are at most $\alpha-1$ requests between each of these pairs.
  By performing a single migration that costs $\alpha$, $\OPT$ may avoid paying the remote serving costs for at most $4 (\alpha - 1)$ requests from $\sigma_E$.
  Thus, for serving $\sigma_E$, $\OPT$ pays at least $W \cdot \frac{\alpha}{4 (\alpha-1)}$.
  This gives us
  \begin{equation*}
    \OPT(\sigma_E) + \OPT(M)  \geq (|\sigma_E|-3(\alpha - 1)\cdot \ell) \cdot \frac{\alpha}{4 (\alpha-1)}
    \geq |\sigma_E| / 4 - 4 \alpha \cdot \ell.
  \end{equation*}
  From that, we obtain $|\sigma_E| \geq 4(\OPT(\sigma_E)+\OPT(M)) + 16\alpha \cdot \ell$.
  Let $E := \OPT(\sigma_E) + \OPT(M)$. Finally, we are ready to bound the competitive ratio
  \begin{equation*}
    \frac{\TAlg(\sigma)}{\OPT(\sigma)} \leq \frac{7\alpha \cdot \ell + |\sigma_E|}{(\alpha + \OPT(\sigma_E) + \OPT(M))/2} \leq \frac{46\alpha\cdot\ell + 8\cdot E}{\alpha + E} \leq 46 \ell = O(\ell).
  \end{equation*}

>>>>

\title{Brief Announcement: Tight Bounds for Online Balanced Repartitioning}

\author{Maciej Pacut}
\email{maciej.pacut@univie.ac.at}
\orcid{0000-0002-6379-1490}
\affiliation{%
  \institution{Faculty of Computer Science, University of Vienna}
  \country{Austria}
}

@@

\keywords{online algorithms, competitive analysis, graph partitioning, clustering}
\acmISBN{}\acmPrice{}
\acmDOI{}

\ccsdesc[500]{Networks~Network algorithms}
\ccsdesc[300]{Computer systems organization~Cloud computing}
\ccsdesc[300]{Computer systems organization~Distributed architectures}

@@

\begin{abstract}
Distributed   applications,  including  batch  processing, streaming, scale-out databases,
or machine learning, generate a significant amount of network traffic. By collocating frequently communicating nodes (e.g., virtual machines) on the same clusters (e.g., server or rack), the network load can be reduced and application performance improved. 
However, as the communication pattern is a priori unknown and may change over time, it needs to be learned efficiently, in an online manner.
%
This paper revisits the online 
balanced repartitioning problem 
(introduced by Avin et al.~at DISC 2016)
which asks for an algorithm that strikes
an optimal tradeoff between the benefits
of collocation (i.e., lower network load) 
and its costs (i.e., migrations). 
%
Our first contribution is a significantly improved
lower bound of $\Omega(k\cdot \ell)$ on the
competitive ratio, where $\ell$ is the number
of clusters and $k$ is the cluster size,
even for a scenario in which the communication
pattern is static and can be perfectly partitioned;
we also provide a tight upper bound 
of $O(k\cdot \ell)$ for this scenario.
In addition, we present a tight upper bound
of $\Theta(\ell)$ for $k=3$,
for the general model in which the
communication pattern can change arbitrarily
over time. 
\end{abstract}

@@

\noindent \textbf{Motivation and model.}
The \emph{balanced repartitioning} problem (\OBRP{})
is a fundamental learning problem
that finds applications in the context of
distributed systems optimization~\cite{repartition-disc}. We are given a set $V$ of $n$ nodes 
(e.g., virtual machines or processes),
initially arbitrarily partitioned into $\ell$~clusters
(e.g., servers or entire racks),
each of size~$k$.
The nodes interact using
a sequence of pairwise communication requests
$\sigma = (u_1,v_1),$ $(u_2,v_2),$ $(u_3,v_3), \ldots$,
where a pair $(u_t,v_t)$ indicates that nodes $u_t$ and $v_t$ exchange a~certain amount of data.
Nodes in $C \subset V$ are \emph{collocated}
if they reside in the same cluster.

An algorithm serves a communication request between two nodes
either \emph{locally} at cost~0
if they are collocated,
or \emph{remotely} at cost~1
if they are located in different clusters.
We refer to these two types of requests as \emph{internal}
and \emph{external} requests, respectively.
Before serving a request,
an online algorithm may perform a \emph{repartition},
%(i.e., \emph{reconfigure}).
i.e.,
it may move (``migrate'') some nodes into clusters different from their current clusters, while respecting the capacity of every cluster. 
Afterwards, 
the algorithm serves the  request.
The cost of migrating a node from one cluster to another
is~$\alpha \in \mathbb{Z}^+$.
For any algorithm $\ALG$,
its cost,
denoted by $\ALG(\sigma)$,
is the total cost of communications and
the cost of migrations performed by $\ALG$ while serving the sequence $\sigma$.

We consider two flavors of the problem
in this paper. In the first one, which
was also studied by Henzinger et al.~\cite{sigmetrics19_partitioning}
(SIGMETRICS 2019) and which
we here mainly consider for the sake of a stronger
lower bound, we assume that $\sigma$
simply reveals the edges of a static graph
that can be perfectly partitioned
(i.e., in principle, no external requests
are required). 
In the second one, we consider a general
model where $\sigma$ can be arbitrary;
this was the original problem introduced
by Avin et al.~\cite{repartition-disc} at DISC 2016.
For simplicity, we will refer to the former
model as the \emph{learning} model (as
one has to learn the static communication graph) 
and to the latter as the \emph{online} model.

\noindent \textbf{Contributions.}
We provide an improved lower bound 
of $\Omega(k\cdot\ell)$ on the competitive ratio of any online deterministic online algorithm 
even in the learning model;
the best known lower bound so far was $\Omega(k)$,
even for a more general model~\cite{repartition-disc}.
We also present an asymptotically optimal, 
$O(k\cdot \ell)$-competitive algorithm
for the restricted model.
For the online model, we present  
an asymptotically optimal,
$\Theta(\ell)$ algorithm for $k=3$;
the best known upper bound 
so far was $O(\ell^2)$~\cite{repartition-disc}.
%
Our algorithms can be distributed
similarly to the approach in~\cite{sigmetrics19_partitioning}.

\noindent \textbf{Related work.}
The existing results in~\cite{repartition-disc}
and~\cite{sigmetrics19_partitioning}
primarily focus on a model where 
the online algorithm can use augmentation,
i.e., has slightly larger clusters than the offline 
algorithm. In contrast, in our paper, we focus
on a scenario where the nodes need to be
perfectly balanced among the clusters.
The problem has also been studied in a weaker
model where the adversary can only sample
requests from a fixed distribution~\cite{stochastic-ring}.
For clusters of size $2$ (i.e., $k=2$), 
it is known that a constant competitive algorithm exists~\cite{repartition-disc}.

%More generally, the model is related to online
%caching~\cite{SleTar85,FKLMSY91,McGSle91,AcChNo00},
%see~\cite{repartition-disc} for a discussion.
%The static offline version of~the~problem, called the
%\emph{$\ell$-balanced graph partitioning problem} is 
%NP-complete, and cannot even be approximated within %any finite factor unless P
%= NP~\cite{AndRae06}.

\noindent \textbf{Preliminaries.}
In this paper, we use the technique of maintaining \emph{(connected) components} of nodes, similar to previous approaches~\cite{repartition-disc}. A~\emph{component} is a subset of frequently communicating nodes.
We use this concept in both algorithms in this paper, but also in the lower bound, to incur high cost for any online algorithm that splits them.
An~algorithm is \emph{component respecting}
if it always keeps nodes that  belong to the same component collocated.
That is,
if the algorithm needs to move a node,
it moves the whole component containing the node.
%We say that the request $(u,v)$ is \emph{external}, if the algorithm is in the configuration, where $u$ and $v$ are located at different clusters.
%This is achieved by opting a sequence of  partitions (a.k.a.~configurations)
%such that frequently communicating nodes (from different clusters)
%are relocated to the same cluster before they inflict too much communication cost.
%A \emph{reconfiguration} (i.e.,\emph{ re-partitioning})
%is performed by migrating nodes between clusters,
%at the total cost of individual node migrations.
%The objective is to jointly minimize communication and migration costs while serving the sequence.

\section{The Learning Problem} %$\Omega(k\cdot \ell)$ for Competitive Ratio of Any Deterministic Algorithm}
\label{sec:lowerbound}

%In this section, we provide a lower bound $\Omega(k\cdot \ell)$ for the competitive ratio of any deterministic online algorithm.
%It is sufficient to have clusters of capacity $k\geq 3$ for this lower bound.
%For $k=2$ a constant competitive algorithm exists.
%At the end of this section, we also
%discuss an upper bound.

\noindent \textbf{Lower bound.}
The adversary constructs components of nodes based on the partitioning choices of a given online algorithm.
Adversary issues requests between pairs of nodes that belong to the same component.
If an online algorithm splits a component at any moment,
the adversary issues requests to the split pair until the algorithm collocates them.
We utilize the fact that the online algorithm does not know the components of the adversary, and it may split some components while evicting nodes.
On the other hand, $\OPT$ initially collocates all nodes of each component and never splits them again.

@@

\begin{theorem}
	The competitive ratio of any deterministic online algorithm for \OBRP{} is in $\Omega(k\cdot \ell)$ for any $k\geq 3$.
\end{theorem}

\begin{proof}
	We construct an instance of the problem with $\ell$ clusters 
	$\set{ S_1, S_2,\dots , S_{\ell}}, |S_i|  = k$.
	Let $I(C)$ denote the cluster where nodes of component $C$ are located initially.
	%	\maciek{So far components were not introduced; the best place to introduce these would be around the definition of \PPOBRP}.
	Fix any online algorithm \ALG{}.
  We join some nodes into components, and produce the input sequence by observing actions of \ALG{} with respect to nodes of these components.
	If \ALG{} at any point splits a component
	(i.e., spreads its nodes over two or more clusters),
	then we continue to issue requests between non-collocated nodes of the component until \ALG{} collocates the nodes of the component.
	If \ALG{} never collocates them, then it is not competitive.

  Initially each node belongs to its own component of size $1$.
  We say that a node is single if it belongs to a singleton component.
  We begin by forming a component of single $k-1$ nodes on the cluster $S_1$.

	Let $x_0$  be the only single node left on $S_1$ and  $y_0 \in S_2$ be any single node on $S_2$.
	Next,
	we merge components of $x_0$~and~$y_0$.
  Since these reside on different clusters of \ALG{}, it produces an external request.
  \ALG{} must collocate them in a~cluster other than $S_1$.
	To this end,
	\ALG{} moves to a new configuration
	that replaces $x_0$ and $y_0$ with two other single nodes $x_1$ and $y_1$ respectively.

	Next,
	we merge the component of $x_1$ and the largest component $C$ s.t.~$I(C) = I(x_1)$.
	\ALG{} must collocate $x_1$ and $C$ in a~cluster other than $S_1$ and
	consequently replace $x_1$ with some other (single) node $x_2$.
	We repeat merging the component of single node $x_i$ on $S_1$ and the largest component $C'$ s.t.~$I(C')=I(x_i)$,
  until there are only two single nodes left that  originate from the same cluster.
	Formally, we denote these two single nodes by $x^*, y^*$, and we have $I(x^*) = I(y^*)$, and~for any other pair of single nodes
	$x'$ and $y'$,
	we have $I(x') \neq I(y')$.
	At this point there are at most $\ell+1$ single nodes left,
	otherwise, there would be more pairs of single nodes that were initially in the same cluster.

	Given this sequence of requests,
	the optimal strategy is to migrate $\set{x_0,y_0}$ to the cluster $I(x^*)$ by
	swapping $\set{x_0,y_0}$ with $\set{x^*,y^*}$.
	Hence,
	OPT pays for $2$ node swaps and
	\ALG{} incurs at least one swap for each node in the sequence $X := x_0, x_1,\dots$.
	We exclude at most $(k-1) + ( \ell+1)$ nodes out of $k \cdot \ell$ nodes,
  therefore $|X| \geq k \cdot \ell - k - \ell$.
  Finally, $\ALG/\OPT \geq (k \cdot \ell - k - \ell) / 2\in \Omega(k\cdot\ell)$.
\end{proof}

\noindent \textbf{Upper bound.}
Now we give an algorithm for a restricted variant of  \OBRP{}, named \PPOBRP{},
%\maciek{This is not sufficient! We also disallow serving requests remotely.}
where an optimal offline algorithm ($\OPT$) moves to a perfect partition
at the beginning and stays there perpetually.
%We refer this variant as \PPOBRP{}.
The task of an online algorithm for \PPOBRP{} is to recover (or learn) the perfect partition while not paying too much relative to \OPT.
%
%\maciek{This is not a valid description of the model. We need the assumption that collocation is needed in our model. Otherwise OPT would just serve some requests remotely, and our analysis would not hold.}
The first request between any two non-collocated nodes reveals an edge of the communication graph.
Our online algorithm collocates them immediately and never separates them since by assumption \OPT does the same.
Revealed edges, up to any point in the sequence,
induce some connected subgraphs which grow as components.
%An algorithm that always  keeps nodes belonging to the same component collocated is \emph{component respecting}.

We assume \OPT begins with the initial configuration
$P_I = I_1, \dots, I_{\ell}$ and moves to the final partitioning
$P_F = F_1, \dots, F_{\ell}$.
 The \emph{distance} of a configuration $P = C_1, \dots, C_{\ell}$ from the initial configuration is the number of nodes in $P$ that do not reside in their initial cluster.
    That is,
    $\dist(P, P_I) := \sum_{j=1}^{\ell} | C_j \setminus I_j |$. 
In other words,
at least $\dist(P, P_I)/2$ node swaps are required in order to reach the configuration $P$ from $P_I$, and thus
$\OPT \geq \Delta:= \dist(P_F, P_I) $.
 With each repartitioning,
  \PPL moves to a configuration that minimizes the distance to the initial configuration $P_I$.
As a result,
\PPL never ends up in a configuration that is more than $\Delta$ (single-node) migrations away from $P_I$.
This invariant ensures that \PPL does not pay too much while recovering $P_F$.
%      \maciek{``re-partition'' procedure name should indicate the fact that this is a specific repartition that is close to initial partition}
      We note that the $\mathit{repartition}$ at Line \ref{line:rebalance} replaces the current configuration $P$ with a~perfect partition closest to $P_I$.
Hence it never moves to a configuration beyond distance $\Delta$.
The scheme of the algorithm can be found in the Appendix \ref{apx:PPL}.  
\begin{property} \label{prop:dist<OPT}
    Let $P$ be any configuration chosen by \PPL at Line $\ref{line:rebalance}$.
    Then, $\dist(P,P_I) \leq \Delta$.
\end{property}

% In Section~\ref{sec:lowerbound} we constructed a $\Omega(k \cdot \ell)$ for \OBRP{}.
% Note that the lower bound holds also in the perfect partition model, as the constructed input sequence allows \OPT to move to a perfect partition.
% The corollary is that PPL is optimal.

\section{The Online Problem}
\label{sec:k3}

Let us now discuss the general online
model where the request sequence
can be arbitrary. We show that the classic \emph{rent-or-buy} approach~\cite{karlin-ski-rental} allows obtaining an optimal algorithm for $k=3$.
Upon receiving $\alpha$ requests between a pair of nodes, we collocate them in one cluster until the end of a phase (that we define precisely later).

%We define the algorithm \TAlg in the following way.
Now we describe the algorithm \TAlg.
It partitions nodes into components, and
initially, each node belongs to its own singleton component.
For each pair of nodes, \TAlg maintains a counter, initiated to $0$. 
Upon receiving a request to a pair that is not collocated in one cluster, it increases their counter by $1$.
If the counter for a pair $(u,v)$ reaches $\alpha$, \TAlg merges the components of $u$ and $v$, and moves to the closest component respecting partitioning.
If no such partitioning exists, \TAlg resets all components to singletons, resets all counters to $0$, and ends the phase.

@@

In our analysis, we distinguish among three types of clusters: $C_1, C_2, C_3$. In a cluster of type $C_i$, the size of the largest component contained in this cluster is $i$.
Before bounding the competitive ratio of \TAlg, we introduce the lemma that estimates the cost of a single repartition of \TAlg. We defer its proof to Appendix~\ref{apx:omitted}.

\begin{lemma}
  \label{lem:1req}
  During a single repartition, \TAlg exchanges at most $2$ pairs of nodes.
\end{lemma}

@@

  As \TAlg is component respecting, it never increases any counter above $\alpha$.
  We say that the pair $(u, v)$ is \emph{saturated} if the counter has value $\alpha$, and \emph{unsaturated} otherwise.
  Let $\sigma$ be the input sequence that arrived during the phase.
  Let $\sigma_{cost}$ be the requests that at the moment of arrival were external requests for \TAlg (these are the only requests that incurred a cost for \TAlg).
  In our analysis, we partition $\sigma_{cost}$ into subsequences $\sigma_I$ and $\sigma_E$.
  The sequence $\sigma_I$ (inter-component requests) are the requests from $\sigma_{cost}$ issued to pairs that belong to the same component of \TAlg at the end of the phase.
  The sequence $\sigma_E$ (extra-component requests) denotes the requests from $\sigma_{cost}$ that do not appear in $\sigma_I$.

  %Let $A^I$ be the cost of (extra-cluster) communication incurred in this phase by \TAlg between pairs that belonged to the single component at the end of the phase.
  %Let $A^E$ be the cost of (extra-cluster) communication incurred in this phase between the nodes that belong to different components at the end of this phase.
  Let $\TAlg(M)$ be the cost of migrations performed by \TAlg in this phase.
  \TAlg performs at most $2 \ell$ component merge operations, as
  exceeding this number means that a component of size $4$ exists, and the phase should have ended already.
  Combining this with Lemma~\ref{lem:1req} gives us $\TAlg(M) \leq 8\alpha\cdot\ell$ (recall that an exchange costs $2\alpha$).
  %Together with Lemma~\ref{lem:1req}, this allows us to bound the cost of migrations, $\TAlg(M) \leq 6\cdot\alpha\cdot\ell$.

  Now we bound $\TAlg(\sigma_I)$.
  A~cluster of type $C_3$ contributes at most $3 \alpha - 1$ to $\TAlg(\sigma_I)$, as $2$ of pairs of nodes from the component are saturated and contribute $\alpha$ each, and the third, unsaturated pair contributes at most $\alpha-1$.
  Other cluster types contribute less: $C_1$ contributes $0$ and $C_2$ contributes $\alpha$.
  Summing this over all $\ell$ clusters gives us $\TAlg(\sigma_I) \leq (3 \alpha-1)\cdot \ell \leq 3\alpha\cdot\ell$.

@@

  Now we estimate the cost of $\OPT(\sigma_E)$ while accounting savings from migrations.
  By performing a swap of nodes $(u,v)$, $\OPT$ collocates $u$ with two nodes $u', u''$, and $v$ with two nodes $v'$, $v''$.
  This may allow serving requests between $(u,u')$, $(u,u'')$, $(v,v')$ and $(v,v'')$ for free afterwards.
  As $\sigma_E$ consists of requests to unsaturated pairs, and it accounts only for external requests, there are at most $\alpha-1$ requests between each of these pairs.
  By performing a single swap that costs $2\alpha$, $\OPT$ may avoid paying the remote serving costs for at most $4 (\alpha - 1)$ requests from $\sigma_E$.
  Thus, for serving $\sigma_E$, $\OPT$ pays at least $\OPT(\sigma_E) + \OPT(M) \geq W \cdot \frac{2\alpha}{4 (\alpha-1)}\geq |\sigma_E| / 2 - 2 \alpha \cdot \ell$.
  From that, we obtain $|\sigma_E| \geq 2(\OPT(\sigma_E)+\OPT(M)) + 4\alpha \cdot \ell$.
  Let $E := \OPT(\sigma_E) + \OPT(M)$. Finally, we are ready to bound the competitive ratio
  \begin{equation*}
    \frac{\TAlg(\sigma)}{\OPT(\sigma)} \leq \frac{11\alpha \cdot \ell + |\sigma_E|}{\alpha/2 + E/2} \leq \frac{27\alpha\cdot\ell + 2\cdot E}{\alpha + E} \leq 27 \ell = O(\ell).
  \end{equation*}
